---
title: "fpp3"
author: "Elad Danovich"
date: "2024-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Time series graphics

```{r}
library(fpp3)
```
## tsibble objects
```{r}
mydata <- tsibble(
  Year = 2015:2019,
  y = c(123, 39, 78, 52, 110),
  index = Year)
mydata
```
Or, we can transform tibble to tsibble
```{r}
mydata <- tibble(
  year = 2015:2019,
  y = c(123, 39, 78, 52, 110)
) |> as_tsibble(index = year)
mydata
```

```{r}
PBS
```
```{r}
prison <- readr::read_csv("https://OTexts.com/fpp3/extrafiles/prison_population.csv")
prison
```
WE get a tibble! Let's transform it to tsibble, first lets ass a quarter column:
```{r}
prison <- prison |>
  mutate(Quarter = yearquarter(Date))
prison
```
We wont need the Date column:
```{r}
prison <- prison |>
  mutate(Quarter = yearquarter(Date)) |>
  select(-Date)
prison
```
Lastly, tranform it to tsibble:
```{r}
prison <- prison |>
  mutate(Quarter = yearquarter(Date)) |>
  select(-Date) |>
  as_tsibble(key = c(State, Gender, Legal, Indigenous),
             index = Quarter)
prison
```

```{r}
PBS
```


```{r}
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(TotalC = sum(Cost)) |>
  mutate(Cost = TotalC / 1e6) -> a10
a10
```



## Time plots

autoplot detects the a10 its a tsibble object so it creates a timeplot
```{r}
a10 |> 
  autoplot(Cost) +
  labs(y = "$ (millions)", title = "Australin antidiabetic drug sales") # + geom_point()
```
```{r}
ansett
```

```{r}
ansett |>
  autoplot(Passengers)
```
Too messy...

```{r}
melsyd_economy <- ansett |>
  filter(Airports == "MEL-SYD", Class == "Economy") |>
  mutate(Passengers = Passengers/1000)

autoplot(melsyd_economy, Passengers) +
  labs(title = "Ansett airlines economy class",
       subtitle = "Melbourne-Sydney",
       y = "Passengers ('000)")
```



## Time series patterns
```{r}
aus_production
```

```{r}
aus_production |>
  filter(year(Quarter) >= 1980) |>
  autoplot(Electricity) +
  labs(y="GWh", title = "Australian electricity production")
```
trend comp - long term increase
seasonal comp - regular peaks every four obs

```{r}
aus_production |>
  autoplot(Bricks) +
  labs(y = "million units", title = "Australian clay brick production")
```
trend comp - mid50s to mid70s increase
seasonal comp - there are peaks in summer quarters


```{r}
us_employment
```
```{r}
us_employment |>
  filter(Title == "Retail Trade", year(Month) >= 1980) |>
  autoplot(Employed / 1e3) + 
  labs(y = "MIllion people", title = "Retail employment, USA")
```
trend comp - long increasing trend
seasonal comp - spikes in the summer months
cyclic comp - periods on increasing and then slow downs

```{r}
gafa_stock
```
```{r}
gafa_stock |>
  filter(Symbol == "AMZN", year(Date) >= 2018) |>
  autoplot(Close) + 
  labs(y = "$US", title = "Amazon closing stock price")
```
Wandering behavior

```{r}
pelt
```
```{r}
pelt |>
  autoplot(Lynx) + 
  labs(y = "Number trapped", title = "Annual canadian Lynx Trappings")
```
cyclic pattern

Differences between seasonal and cyclic patterns:
- seasonal patern constant length; cyclic pattern variable length
- average length of cycle longer than length of seasonal pattern
- magnitude of cycle more variable than magnitude of seasonal pattern



## Seasonal plots
```{r}
a10 |> gg_season(Cost, labels = "both") +
  labs(y = "$ million", title = "Seasonal plot: antidiabetic drug sales")
```
```{r}
beer <- aus_production |>
  select(Quarter, Beer) |> filter(year(Quarter) >= 1992)
beer |> autoplot(Beer) +  geom_point() +
  labs(title = "Australian beer production", y= "Megaliters")
```
Spikes in Q4 - beers for summer!
```{r}
beer |> gg_season(Beer, labels = "right")
```
Now it's more clearly that the spikes in Q4. 
Also we can see that there is a down trend.

```{r}
vic_elec
```
```{r}
vic_elec |> autoplot()
```
```{r}
vic_elec |> gg_season(Demand)
```
Spikes in Australia's summer months, and little higher demnd in winter time
```{r}
vic_elec |> gg_season(Demand, period = "week")
```
Weekend demand < weekday demand
2 spikes in each day(morning and evening)
```{r}
vic_elec |> gg_season(Demand, period = "day")
```
low demand when sleeping.
spike when waking up and getting back home from work.



## Seasonal subseries plots


































## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
