---
title: "02-Time-series-graphics"
author: "Elad Danovich"
date: "2024-05-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Time series graphics

```{r}
library(fpp3)
```

## tsibble objects
```{r}
mydata <- tsibble(
  Year = 2015:2019,
  y = c(123, 39, 78, 52, 110),
  index = Year)
mydata
```
Or, we can transform tibble to tsibble
```{r}
mydata <- tibble(
  year = 2015:2019,
  y = c(123, 39, 78, 52, 110)
) |> as_tsibble(index = year)
mydata
```

```{r}
PBS
```
```{r}
prison <- readr::read_csv("https://OTexts.com/fpp3/extrafiles/prison_population.csv")
prison
```
WE get a tibble! Let's transform it to tsibble, first lets ass a quarter column:
```{r}
prison <- prison |>
  mutate(Quarter = yearquarter(Date))
prison
```
We wont need the Date column:
```{r}
prison <- prison |>
  mutate(Quarter = yearquarter(Date)) |>
  select(-Date)
prison
```
Lastly, tranform it to tsibble:
```{r}
prison <- prison |>
  mutate(Quarter = yearquarter(Date)) |>
  select(-Date) |>
  as_tsibble(key = c(State, Gender, Legal, Indigenous),
             index = Quarter)
prison
```

```{r}
PBS
```


```{r}
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(TotalC = sum(Cost)) |>
  mutate(Cost = TotalC / 1e6) -> a10
a10
```



## Time plots

autoplot detects the a10 its a tsibble object so it creates a timeplot
```{r}
a10 |> 
  autoplot(Cost) +
  labs(y = "$ (millions)", title = "Australin antidiabetic drug sales") # + geom_point()
```
```{r}
ansett
```

```{r}
ansett |>
  autoplot(Passengers)
```
Too messy...

```{r}
melsyd_economy <- ansett |>
  filter(Airports == "MEL-SYD", Class == "Economy") |>
  mutate(Passengers = Passengers/1000)

autoplot(melsyd_economy, Passengers) +
  labs(title = "Ansett airlines economy class",
       subtitle = "Melbourne-Sydney",
       y = "Passengers ('000)")
```



## Time series patterns
```{r}
aus_production
```

```{r}
aus_production |>
  filter(year(Quarter) >= 1980) |>
  autoplot(Electricity) +
  labs(y="GWh", title = "Australian electricity production")
```
trend comp - long term increase
seasonal comp - regular peaks every four obs

```{r}
aus_production |>
  autoplot(Bricks) +
  labs(y = "million units", title = "Australian clay brick production")
```
trend comp - mid50s to mid70s increase
seasonal comp - there are peaks in summer quarters


```{r}
us_employment
```
```{r}
us_employment |>
  filter(Title == "Retail Trade", year(Month) >= 1980) |>
  autoplot(Employed / 1e3) + 
  labs(y = "MIllion people", title = "Retail employment, USA")
```
trend comp - long increasing trend
seasonal comp - spikes in the summer months
cyclic comp - periods on increasing and then slow downs

```{r}
gafa_stock
```
```{r}
gafa_stock |>
  filter(Symbol == "AMZN", year(Date) >= 2018) |>
  autoplot(Close) + 
  labs(y = "$US", title = "Amazon closing stock price")
```
Wandering behavior

```{r}
pelt
```
```{r}
pelt |>
  autoplot(Lynx) + 
  labs(y = "Number trapped", title = "Annual canadian Lynx Trappings")
```
cyclic pattern

Differences between seasonal and cyclic patterns:
- seasonal patern constant length; cyclic pattern variable length
- average length of cycle longer than length of seasonal pattern
- magnitude of cycle more variable than magnitude of seasonal pattern



## Seasonal plots
```{r}
a10 |> gg_season(Cost, labels = "both") +
  labs(y = "$ million", title = "Seasonal plot: antidiabetic drug sales")
```
```{r}
beer <- aus_production |>
  select(Quarter, Beer) |> filter(year(Quarter) >= 1992)
beer |> autoplot(Beer) +  geom_point() +
  labs(title = "Australian beer production", y= "Megaliters")
```
Spikes in Q4 - beers for summer!
```{r}
beer |> gg_season(Beer, labels = "right")
```
Now it's more clearly that the spikes in Q4. 
Also we can see that there is a down trend.

```{r}
vic_elec
```
```{r}
vic_elec |> autoplot()
```
```{r}
vic_elec |> gg_season(Demand)
```
Spikes in Australia's summer months, and little higher demnd in winter time
```{r}
vic_elec |> gg_season(Demand, period = "week")
```
Weekend demand < weekday demand
2 spikes in each day(morning and evening)
```{r}
vic_elec |> gg_season(Demand, period = "day")
```
low demand when sleeping.
spike when waking up and getting back home from work.



## Seasonal subseries plots
```{r}
a10 |>
  gg_subseries(Cost) + 
  labs(y = "$ million", title = "Subseries plot: antidiabetic drug sales")
```
blue line is the average
clear up trend

```{r}
beer |> gg_subseries(Beer)
```
Now it's clear that Q4 is highest and Q2 is the lowet
the decrease in the trend is due to Q4

```{r}
holidays <- tourism |>
  filter(Purpose == "Holiday") |>
  group_by(State) |>
  summarise(Trips = sum(Trips))
holidays
```
```{r}
holidays |> autoplot(Trips) + 
  labs(y = "thousands of trips", title = "Australian domestic holiday")
```
```{r}
holidays |> gg_season(Trips) +
  facet_wrap(vars(State), nrow = 2, scales = "free_y") +
  labs(y = "thousands of trips", title = "Australian domestic holiday")
```
```{r}
holidays |>
  gg_subseries(Trips) +
  labs(y = "thousands of trips", title = "Australian domestic holiday")
```



## Scatterplots
```{r}
vic_elec_day_type <-  vic_elec |>
  filter(year(Time) == 2014) |>
  mutate(Day_Type = case_when(
    Holiday ~ "Holiday",
    wday(Date) %in% 2:6 ~ "Weekday",
    TRUE ~ "Weekend"))
vic_elec_day_type
```
```{r}
vic_elec_day_type |> autoplot(Demand)
```
```{r}
vic_elec_day_type |> autoplot(Temperature)
```
```{r}
vic_elec_day_type |>
  ggplot(aes(x = Temperature, y = Demand, colour = Day_Type)) +
  geom_point() +
  labs(x = "Temperature (C)", y = "Electricity demand (GW)")
```
Non linear relationship
increase when colder and when warmer
the shape is the same for weekdays and weekend
higher demand in weekdays

```{r}
us_change |> GGally::ggpairs(columns = 2:6)
```



## Lag plots
```{r}
new_production <- aus_production |>
  filter(year(Quarter) >= 1992)
new_production
```
```{r}
new_production |> gg_lag(Beer, geom = "point")
```
Each grpah shows y_t plotted against y_t-k for different values of k.
lag4 and lag8 with strong positive corr since this is quartly seasonality.



## Autocorrelation
The autocorrealtions are the correations associated with the above scatterplot.
r1 - is the correlation of lag1 and so on...
```{r}
new_production |> ACF(Beer, lag_max = 9)
```
Like we said in the scatterplot section lag4 and lag8 with high positive corr
```{r}
new_production |> ACF(Beer, lag_max = 9) |> autoplot()
```
correlogram
cool animation in the video
```{r}
retail <- us_employment |>
  filter(Title == "Retail Trade", year(Month) >= 1980)
retail |> autoplot(Employed)
```
uptrend + seasonal
```{r}
retail |> ACF(Employed, lag_max = 48) |>
  autoplot()
```
all are positive due to the trend
AND a little picks each 12 months - the seasonality!

```{r}
google_2015 <- gafa_stock |>
  filter(Symbol == "GOOG", year(Date) == 2015) |>
  select(Date, Close)
google_2015
```
```{r}
google_2015 |> autoplot(Close)
```
```{r}
google_2015 |>
  ACF(Close, lag_max=100) |> autoplot()
```



## White noise
```{r}
set.seed(30)
wn <-  tsibble(t = 1:50, y = rnorm(50), index = t)
wn |> autoplot(y)
```
```{r}
wn |> ACF(y) |> autoplot()
```
```{r}
pigs <- aus_livestock |>
  filter(State == "Victoria", Animal =="Pigs", year(Month) >= 2014)
pigs |> autoplot(Count / 1e3) +
  labs(y = "Thousands", title = "Number of pigs slaughtered in Victoria")
```
test for white noise
```{r}
pigs |> ACF(Count) |> autoplot()
```
2 spikes outside the range, indicate some slight seasonality for lag 2 and 12
meaning, the series is NOT a white noise series.



## Exercises
1.
```{r}
aus_production |> autoplot(Bricks)
```
3.
```{r}
tute1 <- readr::read_csv("/home/elad/git/time-series/fpp3/data/tute1.csv")
tute1
```
```{r}
mytimeseries <- tute1 |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(index = Quarter)
```
```{r}
mytimeseries |>
  pivot_longer(-Quarter) |>
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")
```
5.
```{r}
# install.packages("USgas")
library(USgas)
us_total
```
```{r}
us_total_tsibble <- us_total |> as_tsibble(index = year, key = state)
us_total_tsibble
```
```{r}
us_total_tsibble |> 
  filter(state %in% c("Maine", "Vermont", "New Hampshire", "Massachusetts", "Connecticut", "Rhode Island")) |>
  autoplot(y)
```
6.
```{r}
tourism2 <- readxl::read_excel("/home/elad/git/time-series/fpp3/data/tourism.xlsx")
tourism2
```
```{r}
tourism2 |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(index = Quarter, key = c(Region,State,Purpose))
```

9.
3-D -- up trend + peak in lag12
2-A -- high corr each ~12 months and low corr each ~6 months (probably more accidents in winter time)
4-C -- every 10 years there is a low so hih corr in 10
1-B -- ???
































## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
