---
title: "04-Time-series-features"
author: "Elad Danovich"
date: "2024-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Time series features

```{r}
library(fpp3)
```


## Some simple tatistics
```{r}
tourism |>
  features(Trips, list(mean = mean)) |>
  arrange(mean)
```
```{r}
tourism |> features(Trips, quantile)
```



## ACF Features
```{r}
tourism |> features(Trips, feat_acf)
```
the first autocorrelation coefficient from the original data;
the sum of squares of the first ten autocorrelation coefficients from the original data;
the first autocorrelation coefficient from the differenced data;
the sum of squares of the first ten autocorrelation coefficients from the differenced data;
the first autocorrelation coefficient from the twice differenced data;
the sum of squares of the first ten autocorrelation coefficients from the twice differenced data;
For seasonal data, the autocorrelation coefficient at the first seasonal lag is also returned.



## STL Features




























## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
