---
title: "05-The-forecaster's-toolbox"
author: "Elad Danovich"
date: "2024-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 05-The-forecaster's-toolbox
```{r}
library(fpp3)
```

## A tidy forecasting workflow
1. Preparing data
2. Data visualization
3. Specifying a model
4. Model estimation
5. Accuracy and performance evaluation
6. Producing forecasts
```{r}
gdppc <-  global_economy |>
  mutate(GDP_per_capita = GDP / Population) |>
  select(Year, Country, GDP, Population, GDP_per_capita)
gdppc
```

```{r}
gdppc |>
  filter(Country == "Sweden") |>
  autoplot(GDP_per_capita) +
  labs(title = "GDP per capita for sweden", y = "$US")
```

```{r}
fit <-  gdppc |>
  model(trend_model = TSLM(GDP_per_capita ~ trend()))
fit
```
mabel = model table, each cell corresponds to fitted model.

```{r}
fit |> forecast(h = "3 years")
```
fable = forecast table with point forecasts and distributions.

```{r}
fit |>
  forecast(h = "3 years") |>
  filter(Country == "Sweden") |>
  autoplot(gdppc) + labs(title = "GDP per capita for Sweden", y = "$US")
```



## Some simple forecasting methods
1. forecast the average (mean(y))
2. forecast the last observed value (naive(y))
3. forecast equal to last value from same season (snaive, y ~ lag(m))
4. forecast equal to last value plus average change (rw, y  ~ drift())

```{r}
brick_fit <-  aus_production |>
  filter(!is.na(Bricks)) |>
  model(
    Seasonal.naive = SNAIVE(Bricks),
    Naive = NAIVE(Bricks),
    Drift = RW(Bricks ~ drift()),
    Mean = MEAN(Bricks)
  )
brick_fit
```

```{r}
bricks_fc <-  brick_fit |>
  forecast(h = "5 years")
bricks_fc
```

```{r}
bricks_fc |>
  autoplot(aus_production, level = NULL) +
  labs(title = "Clay brick prduction in Australia", y = "Millions of Bricks") +
  guides(colour = guide_legend(title = "Forecast"))
```

```{r}
# Extract training data
fb_stock <-  gafa_stock |>
  filter(Symbol == "FB") |>
  mutate(trading_day = row_number()) |>
  update_tsibble(index = trading_day, regular = TRUE) #since there are days that the market is close the naive model will pop an error, for example for monday there                                                            isnt a data from sunday to forecast

# Specify, estimate and forecast
fb_stock |>
  model(
    Mean = MEAN(Close),
    Naive = NAIVE(Close),
    Drift = RW(Close ~ drift())
  ) |>
  forecast(h = 42) |>
  autoplot(fb_stock, level = NULL) +
  labs(title = "Facebook closing stock price", y = "$US") + 
  guides(colour = guide_legend(title = "Forecas"))
```








## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
