---
title: "05-The-forecaster's-toolbox"
author: "Elad Danovich"
date: "2024-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 05-The-forecaster's-toolbox
```{r}
library(fpp3)
```

## A tidy forecasting workflow
1. Preparing data
2. Data visualization
3. Specifying a model
4. Model estimation
5. Accuracy and performance evaluation
6. Producing forecasts
```{r}
gdppc <-  global_economy |>
  mutate(GDP_per_capita = GDP / Population) |>
  select(Year, Country, GDP, Population, GDP_per_capita)
gdppc
```

```{r}
gdppc |>
  filter(Country == "Sweden") |>
  autoplot(GDP_per_capita) +
  labs(title = "GDP per capita for sweden", y = "$US")
```

```{r}
fit <-  gdppc |>
  model(trend_model = TSLM(GDP_per_capita ~ trend()))
fit
```
mabel = model table, each cell corresponds to fitted model.

```{r}
fit |> forecast(h = "3 years")
```
fable = forecast table with point forecasts and distributions.

```{r}
fit |>
  forecast(h = "3 years") |>
  filter(Country == "Sweden") |>
  autoplot(gdppc) + labs(title = "GDP per capita for Sweden", y = "$US")
```


















## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
